Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-06-22 Paper/paper.tex
added description of fig.6,7

We construct NSBH waveforms using the LEA+ model. We inject these into zero
noise and run MCMC on the injections using BBH templates (LEA+ with tidal terms
set to zero). We perform this analysis for a variety of mass ratios
$q\in\{2,3,4,5\}$, i.e. $\mbh\in\{2.7M_\odot, 4.05M_\odot, 5.4M_\odot, 6.75M_\odot\}$,
and BH spins $\chibh\in\{-0.5, 0, +0.5, +0.75\}$. We always use $\mns=1.35M_\odot$ and
$\chins=0$, but explore $\lambdans\in\{500, 800, 1000, 1500, 2000\}$. The signal
strength is varied from $\rho=20\rightarrow70$. Our choice of this SNR range is
motivated by previous studies which suggest that signatures of NS tidal effects
become observable only at relatively large SNRs~\cite{Lackey:2013axa}, and is validated
by the results to follow. At design sensitivity, if we expect $10-300$ NSBH detections 
a year~\cite{Abadie:2010cfa}, and if BH masses are uniformly likely up to
$\sim 35M_\odot$~\cite{LIGOVirgo2016a}, we would expect to detect $0.3-7$ disruptive
NSBH mergers a year with $\rho\geq 20$, $0.07-2$ a year with $\rho\geq 30$, and $0.04-1$ a
year with $\rho\geq 70$.

@@

\includegraphics[width=0.65\columnwidth]{plots/LambdaMc2D_q4_mc2_25_chi0_50_snr30}
\includegraphics[width=0.65\columnwidth]{plots/LambdaEta2D_q4_mc2_25_chi0_50_snr30}
\includegraphics[width=0.65\columnwidth]{plots/LambdaChiBH2D_q4_mc2_25_chi0_50_snr30}
\caption{\prayush{INCOMPLETE..}We illustrate here two-dimensional probability distributions, showing
the correlation between the NS tidal deformability parameter $\lambdans$ 
and binary mass and spin parameters.

@@

In the previous section we showed that we begin to care for the effects of NS's
deformation due to its companion hole's tidal field, if (a) the BH is sufficiently
small, (b) BH spin is aligned and $\gtrsim +0.4$, and (c) the NS is
not very compact, with $\lambdans\gtrsim 1000$. The first two conditions
enhance tidal effects on emitted GWs by increasing the number of orbits at small
separation, where the deviation of NSBH and BHBH dynamics is the largest. The third,
i.e. (c), simply makes the NS more easily deformable. All of them reduce the onset 
frequency for NS's disruption close to merger. Therefore, systems that fulfill the 
above conditions have considerably (or, at least, the most) different GW morphology
when compared to equivalent BHBH coalescences. Ultimately, the extent of 
tidal modulations of emitted GWs depends on the equation of state of neutron star
matter, which becomes an indirect measurable in GW astrophysics.
% 
However, if we do indeed receive GWs from such a binary source, what information can we
possibly extract from it about the internal structure of neutron stars?
In this section we consider this question in perspective of measuring the tidal 
deformability parameter of neutron stars purely from gravitational waves they emits 
while orbiting spinning black holes.
%

  `
% \red{Is this the LEA+ mode? If yes, say so}
% \prayush{%
We sample the same set of systems as in the previous section, i.e. those with
$q=\{2,3,4,5\}$, $\chibh=\{-0.5,0,+0.5,+0.75\}$, and
$\lambdans=\{500, 800, 1000, 1500, 2000\}$. For each unique combination of
these parameters, we generate separate injections at SNRs
$\rho=\{20,30,50,70\}$, using the LEA+ model~\cite{Lackey:2013axa}. We perform a
full Bayesian analyses on each of these injections using LEA+ templates, and obtain
probabilistic estimates for source binary parameters from the GW information
alone, including for the NS tidal deformability parameter $\lambdans$.

\red{Present posterior distributions of selected configurations first..}

\red{%
\begin{enumerate}
 \item State the prior on $\lambdans$ because $\lambdans$ weakly deforms waveforms,
I would expect that measurement uncertainty in $\lambdans$ is approx. independent
of $\lambdans$, except for running into the $\lambdans\geq 0$ prior.
 \item Is $\lambdans$ measurement degenerate with other parameters? Either cite 
 known studies [if any], or provide 2-d posterior plots.
\end{enumerate}
}
The precision with which $\lambdans$ can be measured is given by the width of 
its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.

>>>>

We construct NSBH waveforms using the LEA+ model. We inject these into zero
noise and run MCMC on the injections using BBH templates (LEA+ with tidal terms
set to zero). We perform this analysis for a variety of mass ratios
$q\in\{2,3,4,5\}$, i.e. $\mbh\in\{2.7M_\odot, 4.05M_\odot, 5.4M_\odot, 6.75M_\odot\}$,
and BH spins $\chibh\in\{-0.5, 0, +0.5, +0.75\}$. We always use $\mns=1.35M_\odot$ and
$\chins=0$, but explore $\lambdans\in\{500, 800, 1000, 1500, 2000\}$. The signal
strength is varied from $\rho=20\rightarrow70$. Our choice of this SNR range is
motivated by previous studies which suggest that signatures of NS tidal effects
become observable only at relatively large SNRs~\cite{Lackey:2013axa}, and is validated
by results that follow. At design sensitivity, if we expect $10-300$ NSBH detections 
a year~\cite{Abadie:2010cfa}, and if BH masses are uniformly likely up to
$\sim 35M_\odot$~\cite{LIGOVirgo2016a}, we would expect to detect $0.3-7$ disruptive
NSBH mergers a year with $\rho\geq 20$, $0.07-2$ a year with $\rho\geq 30$, and $0.04-1$ a
year with $\rho\geq 70$. For our priors, we choose uniform distribution for 
both component masses as well as for black hole spins, i.e. $\mbh\in[2,20]M_\odot$,
$\mns\in[1,3]M_\odot$ and $-0.5\leq \chibh\leq +0.75$. \prayush{FIXME}

@@

\includegraphics[width=0.5\columnwidth]{plots/McEta2D_q4_mc2_25_chi0_50_snr30}
\includegraphics[width=0.5\columnwidth]{plots/EtaChiBH2D_q4_mc2_25_chi0_50_snr30}
\includegraphics[width=0.5\columnwidth]{plots/McChiBH2D_q4_mc2_25_chi0_50_snr30}\\
\includegraphics[width=0.5\columnwidth]{plots/LambdaMc2D_q4_mc2_25_chi0_50_snr30}
\includegraphics[width=0.5\columnwidth]{plots/LambdaEta2D_q4_mc2_25_chi0_50_snr30}
\includegraphics[width=0.5\columnwidth]{plots/LambdaChiBH2D_q4_mc2_25_chi0_50_snr30}\\
\includegraphics[width=0.47\columnwidth]{plots/LambdaMc2D_q4_mc2_25_chi0_50_snr50}
\includegraphics[width=0.47\columnwidth]{plots/LambdaEta2D_q4_mc2_25_chi0_50_snr50}
\includegraphics[width=0.47\columnwidth]{plots/LambdaChiBH2D_q4_mc2_25_chi0_50_snr50}
We illustrate here two-dimensional joint probability distributions, as
recovered for the intrinsic parameters of a binary with
$q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$,
$\lambdans=2000$, and $\rho=30$ (same as in
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}).
The top row shows the same for combinations of 
non-tidal parameters, i.e. binary masses and spins, while the middle row 
shows combinations of $\lambdans$ with non-tidal parameters. The bottom row
is similar to the middle, with the only difference being that the SNR has been
turned up to $\rho=50$.

@@

In the previous section, we showed that the effects of NS's deformation due to
its companion hole's tidal field become discernible in the GW spectrum only when
the following hold: (a) the BH has a sufficiently small mass, (b) BH spin
is positive aligned, i.e. $\chibh\gtrsim +0.4$, (c) the NS is not very compact,
with $\lambdans\gtrsim 1000$, and finally, (d) the source is close enough
such that SNR $\gtrsim 30$.
% 
The first two conditions both enhance the tidal distortion of the star and increase
the number of orbits the system goes through at small separation, where the
differences between NSBH and BHBH signals are maximal.
% 
The first three also reduce the onset frequency of NS's disruption, allowing
for it to happen earlier in the orbit. 
% 
We expect that these conditions are also the ones which would maximize the
likelihood of {\it measuring} tidal effects in NSBH signals. We, therefore,
%
%In addition, a SNR of $30$ corresponds to a $1.35+5M_\odot$ binary with 
%$\chibh = 0.4$ at a luminosity distance of $XXX$~Mpc, if optimally aligned
%with LIGO ($XXX$~Mpc if the binary subtends a $45^o$ angle with LIGO's line of
%sight.
%
%These conditions are strong, but not unrealistic. For e.g., a detection rate
%of $20$ NSBHs a year implies $1-2$ of them will have SNRs$\gtrsim 30$, since
%the number of expected events scales as $1/\rho^2$ if we assume that sources
%are distributed uniformly in spacial volume.
%
turn the question around and ask: under similarly favorable circumstances, can
we gain insights about the internal structure of neutron stars from GW observations.

%
%Ultimately, the extent of tidal moduilations of emitted GWs depends on the
%equation of state of neutron star matter, which becomes an indirect
%measurable in GW astrophysics.
% 
%We therefore turn to the question: If we do observe GWs from disruptive 
%coalescing NSBHs, what information can we extract from them about the internal
%structure of neutron stars?
% 
In order to explore this, we calculate here the accuracy with which we measure
$\lambdans$ from {\it single} GW observations. We sample the same set of
disruptive NSBH systems as in the previous section, i.e. those with
$q=\{2,3,4,5\}$, $\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000, 1500, 2000\}$.
For each unique combination of these parameters, we generate separate injections
at SNRs $\rho=\{20,30,50,70\}$, using the LEA+ model~\cite{Lackey:2013axa}. We
perform full Bayesian analyses on each of these injections using LEA+ templates,
and obtain probabilistic estimates for source binary parameters using GW information
alone (including the NS tidal deformability parameter $\lambdans$). Our priors
on component masses and spins remain as in the previous section, while $\lambdans$
is sampled uniformly from $[0,4000]$.

As an illustration, we show the recovered
probability distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix $q = m_1/m_2 =
5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, while varying $\lambdans$
over $\{1000, 1500, 2000\}$ between the three panels. The SNR is fixed at $\rho=50$.
The true value of $\lambdans$ is given the green (dashed) vertical line,
while the median recovered value is marked by a red (dashed) line. The darker
shaded region marks the $90\%$ confidence interval on $\lambdans$. One thing
to note in this figure is that the measurement uncertainty on $\lambdans$, i.e.
the full width of its $90\%$ confidence interval, grows slower than $\lambdans$
as $\lambdans$ is turned up (while the fractional uncertainty 
$(\Delta\lambdans)^{90\%}/\lambdans$ decreases with increasing $\lambdans$).
%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
$\rho=30$ (instead of $50$). In the top panel, we show the correlation between
non-tidal parameters, as manifested in their recovered joint probability
distributions.
In the middle row, we show the same between $\lambdans$ and non-tidal 
parameters, i.e. chirp mass (left panel), mass ratio $\eta$ (middle panel), and
BH spin $\chibh$ (right). The same is shown in the row below, with the
difference that the SNR has been turned up to $50$ to bring out the intrinsic
ambiguity contours.
%
We find that the measurement of $\lambdans$ is only weakly degenerate with
other parameters, and may not improve much if we knew non-tidal parameters
to better accuracy. The predominant factor that would improve the measurement
accuracy for $\lambdans$ is the signal strength, SNR.

%\red{Present posterior distributions of selected configurations first..}

%\red{%
%\begin{enumerate}
% \item State the prior on $\lambdans$ because $\lambdans$ weakly deforms waveforms,
%I would expect that measurement uncertainty in $\lambdans$ is approx. independent
%of $\lambdans$, except for running into the $\lambdans\geq 0$ prior.
% \item Is $\lambdans$ measurement degenerate with other parameters? Either cite 
% known studies [if any], or provide 2-d posterior plots.
%\end{enumerate}
%}

The precision with which $\lambdans$ can be measured is given by the width of 
its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.
